Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2021-12-01 manu-draft.Rmd
abstract shortening, adding seagrass statement

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen loads in a matter of days.  In the period immediately after the release, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.  Reported fish kills tracked *K. brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions in Tampa Bay.

>>>>

From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA). An estimated 186 metric tons of total nitrogen were exported to the Bay, exceeding typical annual external nitrogen loads in a matter of days. Elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrass resources showed minimal changes over the study period.  By comparing these results to decades of baseline environmental monitoring data, we conclude that many of the biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions.
